---
# title: "Lecture 03"
# author: "Kasthuri Kannan and Arnon Lieber"
# date: '`r format(Sys.Date(), "%b %d, %y")`'
output: html_document
---


### R fundamentals 02: Advanced data types (`r format(Sys.Date(), "%b %d, %y")`)


---

In this lecture we will consider other data types such as **Factors**, **Lists** and **Data Frames** 

---

#### Factors

Factors are determined through categorical variables. What are categorical variables?

* Limited number of differing values.
* Belong to a certain category.
* For statistical analysis, R calls them factors.

##

##### Creating and manipulating factors

```{r eval=TRUE}
  # Create a blood group vector
  blood_group_vector <- c("AB", "O", "B+", "AB-", "O", "AB", "A", "A", "B", "AB-")
  blood_group_vector
```

```{r eval=TRUE}
  # Create fatcors from the vector
  blood_group_factor <- factor(blood_group_vector)
  blood_group_factor
```
  
  
**Note:**

  1. The levels are sorted alphabetically. 
  2. No more quotation marks.

##

R encodes factors to integers for easier memory access and computations. This is done alphabetically. For example, A is assigned 1, AB is assigned 2 etc. This can be viewed by invoking the `str()` function:

```{r eval=TRUE, echo=TRUE}
  # Show the structure of the blood group factor
  str(blood_group_factor)
```
  
<br>
```{r results='hold'}
print(blood_group_vector)
paste(as.character(as.integer(blood_group_factor)), " ")
```


##

This can be over-ridden by specifying the <span style="color:#990000">`levels`</span> argument for the `factor()` function.

```{r eval=TRUE}
  # Define another set of levels over-riding default
  blood_group_factor2 <- factor(blood_group_vector, levels = c("A", "B", "B+", "AB", "AB-", "O"))
  print(blood_group_factor2)
  str(blood_group_factor2)
```

```{r collapse=TRUE}
# Comparing the default alphabetic order with the new one:
as.integer(blood_group_factor)
as.integer(blood_group_factor2)
```

## 

Renaming factors can be done using the `level()` function.

```{r eval=TRUE}
  # Define blood type
  blood_type <- c("B", "A", "AB", "A", "O")
  # Find the factors
  blood_type_factor <- factor(blood_type)
  blood_type_factor
  # Rename the factors
  levels(blood_type_factor) <- c("BT_A", "BT_AB", "BT_B", "BT_O")
  blood_type_factor
```


##

**Note**: It is **extremely important** to follow the **same order** as the default order supplied by R. Otherwise, the result can be extremely confusing as the following exercise will show.



<br>

---

<span style="color:blue">**Classwork/Homework**</span>: Rename the blood_type_factor in the above example as follows: 

<small>`levels(blood_type_factor) <- c("BT_A", "BT_B", "BT_AB", "BT_O")`</small>

and justify the result of outputting `blood_type_factor`. Use `str()` to support your answer.

---

<br>


If you want to safely rename your levels or to change their default order, it is always best to define the `labels` along with the `levels` like this -

```{r eval=FALSE}
  factor(blood_type_factor, levels=c("A", "B", "AB", "O"), 
                            labels=c("BT_A", "BT_B", "BT_AB", "BT_O"))
```

<br>
An easy and fast way to generate a simple factor with given number of repetitions is by the function `gl()`
```{r, collapse=TRUE}
factorZ <- gl(3, 2, length = 12)

print(factorZ)
```


<br>

##### Nominal vs. Ordinal factors

_Nominal factors_: These are categorical variables that cannot be ordered, like blood group. For example, it doesn't make sense to say blood group A < blood group B.

_Ordinal factors_: These are categorical variables that can be ordered. For instance, tumor sizes. We can say T1 (tumor size 2cm or smaller) < T2 (tumor size larger than 2cm but smaller than 5 cm).

##

R provides us with a way to impose order on factors. Simply use the argument `ordered=TRUE` inside the factor function.

```{r eval=TRUE, collapse=F}
  # Specify the tumor size vectore
  tumor_size <- c("T1", "T1", "T2", "T3", "T1")
  # Set the order by specifying "ordered=TRUE"
  tumor_size_factor <- factor(tumor_size, ordered = TRUE,
                              levels=c("T1", "T2", "T3"))
  
  # Print the results
  tumor_size_factor
  
  # Compare one factor vs the other
  tumor_size_factor[1] < tumor_size_factor[2]
```

##

---

<span style="color:blue">**Classwork/Homework**</span>: Use the inequality operator (< or >) on a **nominal** category and print the output.

---

<br>



#### Lists

Recall vectors and matrices can hold only one data type, like integer or character. Lists can hold multiple R objects, without having to perform coercion.

##

```{r eval=TRUE, collapse=T}
  # Defining different data type as vector (Note, coercion takes place)
  vec <- c("Blood-sugar", "High", 140, "mg/dL")
  vec
  
```

```{r eval=TRUE, collapse=T}
  # And as a list
  lst <- list("Blood-sugar", "High", 140, "mg/dL")
  # One can use the is.list() function to see if something is a list
  is.list(lst)
   
  lst
```

##

```{r eval=TRUE}

```

Naming a list can be done through the `names()` function or specifying it in the list itself.

```{r eval=TRUE, collapse=TRUE}
  # Define a list
  lst <- list("Blood sugar", "High", 140, "mg/dL")
  # Assign names and print
  names(lst) <- c("Fluid", "Category", "Value", "Units")
  
  print(lst)
```

##

Or specify names directly while defining the list

```{r eval=TRUE}
  # Specify while constructing the list 
  blood_test <- list(Fluid="Blood sugar", Category="High", Value=140, Units="mg/dL")
  # For compact display use the str() function
  str(blood_test)
```  

##

**Note**: A list can contain another list, or any number of nested lists.

##

##### Aceesing and extending lists

The difference between `[]` and `[[]]` is that, `[]` will return a list back and `[[]]` will return the elements in the list.

```{r eval=TRUE, collapse=TRUE}

  # Creating a list of patient's details containing the 'blood_test' list
  patient <- list(Name="Mike", Age=36, Btest = blood_test)
  
  # Show the first element of the list
  patient[1]
  
  class(patient[1])
  
  # Access the content of the first element
  patient[[1]]
  
  class(patient[[1]])
```
  
##
  
```{r, collapse=TRUE}

  # Show the structure of the third element of the list
  str(patient[3])
  
  # Show the structure of the content of the third element (which in this case is a list by itself)
  str(patient[[3]])
```


<br>

---

<span style="color:blue">**Classwork/Homework**</span>: 

1. What does `patient[c(1,3)]` give us? Is it a list or elements?
2. What does `patient[[c(1,3)]]` give us? Is it a list or elements? 
3. How about `patient[[c(3,1)]]`? What is the difference?  
( _Hint_: `patient[[c(1,3)]]` is same as `patient[[1]][[3]]`).

---

<br>

##

Subsetting by names is super easy: just supply the name within brackets. For example, `patient["Name"]` or `patient[["Name"]]`. 

Subsetting by logicals will work only for returning elements of the list. For instance, `patient[c(TRUE,FALSE)]`. 

It doesn't make sense to return the elements through logicals, i.e., `patient[[c(TRUE,FALSE)]]`.

##

Another cool way to access elements (just the same as using `[[]]`) is the use of `$` sign. 

However, to do this, the list should be named. For example, `patient$Name` will print the patient name.

```{r collapse=TRUE}
class(patient$Name)
```


`$` sign can also be used for extending lists:

```{r eval=FALSE}
  # Extend the list to include gender
  patient$Gender <- "Male"
  # This is same as using double brackets
  patient[["Gender"]] <- "Male"
  # Extend the blood test list to include the date of the test
  patient$Btest$Date <- "Sept.14"
```


##

---

<span style="color:blue">**Classwork/Homework**</span>: How do you remove an element from a list?

---

<br>

#### Data frames

Datasets come with various shapes and sizes. Usually they constitute:

* Observations (eg. each row is an observation)
* Variables (eg. each column is a variable)


Limitations of other data types:

* Matrices consist of only one data type
* Working with lists is not practical 

##

Data frames can contain different values for each observation/row; however, each variable (or a column) should have the same data type. 

Usually data frames are imported - through CSV, or Excel etc. However, we can create a data frame as well.

```{r eval=TRUE}
  # Create name, age and logical vectors
  name <- c("Anne", "James", "Mike", "Betty")
  age <- c(20, 43, 27, 25)
  cancer <- c(TRUE, FALSE, FALSE, TRUE)
  # Form a data frame
  df <- data.frame(name, age, cancer)
  df
```

##

Update the `names` attribute

```{r eval=TRUE}
  # (the same way like we did for vectors)
  names(df) <- c("Name", "Age", "Cancer_Status")
  attributes(df)
  # Or specify directly while creating the data frame
  df <- data.frame(Name=name, Age=age, Cancer_Status=cancer)
  df
```

##

---

<span style="color:blue">**Classwork/Homework**</span>: 

1. Examine the structure of the data frame.
2. What happens if one of the vectors has an unequal length?

---

<br>

##

**Note**: Data frames store character vectors as factors. You can override this as follows: 
```{r, eval=F}
df <- data.frame(Name=name, Age=age, Cancer_Status=cancer, 
                 stringsAsFactors = FALSE)
```
 


##

##### Manipulating data frames: **Subsetting**

```{r eval=TRUE}
  print(df)
```

We can subset by indices:

```{r eval=TRUE, collapse=T}
# Subsetting by indices - works just like matrices
  df[1,2]

# Get the entire row/column - just like matrices
  # Get the second row
  df[2,]

```

##

We can also subset using the names as well as indices:

```{r eval=TRUE, collapse=T}
# Get the "cancer" column
  df[,"Cancer_Status"]  

# One can use column names as well
  df[1, "Age"]

# Get all 2nd and 3rd observation with "name"" and "cancer"" status
  df[c(2,3), c("Name", "Cancer_Status")]
```  

<br><small>
The main difference in subsetting a data.frame versus a matrix is when you specify a single number as index within `[]`. For matrices you get **an element** corresponding to the linear index but for a data frame we'll get the **column vector** that corresponds to the index. </small>

##

An example:
```{r eval=TRUE}
# Print the class (of the values) of the second column
class(df[,2])

# Class of the retrieved element, using a single bracket
class(df[2])
```

<br>
This is because data frames are made up of lists of vectors of equal length. Thus, single `[2]` will correspond to the second element in the list, which is a _vector_ of ages.

##

---

<span style="color:blue">**Classwork/Homework**</span>: Test the operations of lists (like `["age"] & [["age"]]`) on data frames.

---

<br>

##

##### Manipulating data frames: **Extending**

Adding a column is super easy. Since data frames are a list of vectors one can just append a vector to the list. 

For instance, if we have a column of tumor size info like this for each patient: `c("T0","T3","T2","T0")`, the following code will append the vector.

```{r eval=TRUE}
  # Append tumor size vector
  df$Tumor_size <- c("T3", "T0", "T0", "T2")
  df
```

##

<br>
<span style="color:blue">**Classwork/Homework**</span>: 

1. Use `cbind()` to append a vector of choice.
2. What happens if the length of the appending vector is greater than (or less than) row dimensions?

---

<br>

##

In contrast, extending a row (or observation) is not straight-forward. This is because observations may contain different data types. To add observations, make a new data frame and append:

```{r eval=TRUE}
  # Create a data frame (pay attention to the capital letters at the variable names!)
  tom <- data.frame(Name="Tom", Age=47, Cancer_Status="TRUE", Tumor_size="T2")
  # And append
  df <- rbind(df, tom)
  df
```  

---

<span style="color:blue">**Classwork/Homework**</span>:

1. Can you use the `list()` function instead of the data frame function in the above code?
2. What happens if you leave the arguments `name=`, `age=` etc. in the above code?
3. Check out the function `expand.grid()`, what is it for?  
   Try out the following example and think when it is useful.


```{r eval=F}
expand.grid(height = as.character(seq(60, 70, 5)), weight = seq(100, 200, 50),
            sex = c("Male","Female"), stringsAsFactors = FALSE)
```

---

<br>

##### Manipulating data frames: **Sorting**

We can use the `order()` function to sort the entire data frame with respect to a particular column.

```{r eval=TRUE, collapse=TRUE}
  # Rank the entries of a column, say "Age"
  ranks <- order(df$Age)
  
  # `ranks` is a vector of indexes
  print(ranks)
  
  # Sort the data frame according to the rank
  df[ranks,]
  
```

##

---

<span style="color:blue">**Classwork/Homework**</span>:

1. Why does `sort(df$age)` return an error?
2. When you fix the problem, what does the command perform and how is it related to the ranks?
3. Sort the entries in descending order of the age. _(Hint: Question the function to find out more about the function in question)_.


<br>

#### Selected materials and references

[An Introduction to R - Chapter 6](https://cran.r-project.org/doc/manuals/R-intro.pdf)  
[R Tutorial-List](http://www.r-tutor.com/r-introduction/list)  

---

In the next lecture we will learn about **graphics** in R.


---
  

